% Remove chapter numbering for CV
\chapter*{Curriculum Vitae}
\addcontentsline{toc}{chapter}{Curriculum Vitae}
% Force single spacing and remove all extra spacing
\begin{singlespace}
\footnotesize % Make font smaller for CV
\setlength{\parindent}{0pt} % no indent for CV
\setlength{\parskip}{0pt}   % no space between paragraphs
\setlength{\itemsep}{0pt}   % no space between items
\setlength{\parsep}{0pt}    % no space between paragraphs in lists
\setlength{\topsep}{0pt}    % no space above/below lists
\setlength{\partopsep}{0pt} % no extra space above/below lists

\subsection*{Daniel Kadyrov}

Email: \href{mailto:dkadyrov@stevens.edu}{dkadyrov@stevens.edu}\\

\subsection*{Education}
\textbf{Stevens Institute of Technology} \hfill Hoboken, NJ \\
Ph.D. Ocean Engineering \hfill Expected Spring 2026 \\
M.S. Computer Science (Machine Learning and Data Science) \hfill May 2021 \\

\textbf{Binghamton University, State University of New York} \hfill Binghamton, NY \\
M.S. Mechanical Engineering (Acoustics and Dynamic Systems) \hfill May 2016 \\
B.S. Mechanical Engineering \hfill May 2015 \\

\subsection*{Professional Experience}

\textbf{Department of Homeland Security}\\
National Urban Security Technology Laboratory (NUSTL) \hfill New York, NY \\
\textit{General Engineer \hfill May 2019 - Present} \\

\textbf{Stevens Institute of Technology}\\
Sensor Technology and Applied Research Center \hfill Hoboken, NJ \\
\textit{Research Engineer \hfill May 2019 - December 2023} \\

\textbf{Thornton Tomasetti}\\
Weidlinger Applied Science \hfill New York, NY \\
\textit{Senior Engineer \hfill January 2017 - November 2018} \\

\section*{Teaching Experience}
\textbf{Pace University} \hfill New York, NY \\
\textit{Adjunct Professor \hfill September 2023 - Present} \\

\textbf{Columbia University} \hfill New York, NY \\
\textit{Adjunct Lecturer \hfill June 2023 - August 2024} \\

\subsection*{Journal Publications}
\begin{itemize}
  \item D. Kadyrov, A. Sutin, N. Sedunov, A. Sedunov, and H. Salloum, “Acoustic Detection of Insects in Stored Products in the Presence of Strong Ambient Noise,” Sensors, Feb. 2026, doi: 10.3390/s26051511.
  \item D. Kadyrov, A. Sutin, N. Sedunov, A. Sedunov, and H. Salloum, “Vibro-Acoustic Signatures of Various Insects in Stored Products,” Sensors, vol. 24, no. 20, Oct. 2024, doi: 10.3390/s24206736.
\end{itemize}

\subsection*{Conferences}
\begin{itemize}
  \item DHS Research Development \& Testing and Evaluation Summit (Online, 2025)\\
  \textit{Stored Product Insect Detection Systems (SPIDS)}
  \item Artificial Intelligence Meets Science (AIMS) Conference (New York, 2024)\\
  \textit{Perspectives of Machine Learning for Acoustic Methods of Stored Product Insect Detection}
  \item Acoustic Society of America (Denver, 2022)\\
  \textit{D. Kadyrov, A. Sedunov, N. Sedunov, A. Sutin, H. Salloum, and S. Tsyuryupa, “Improvements to the Stevens drone acoustic detection system,” Proc. Mtgs. Acoust., vol. 46, no. 1, p. 045001, Aug. 2022, doi: 10.1121/2.0001602.}
\end{itemize}
\end{singlespace}

